Constraints on Cosmic Rays, Magnetic Fields, and Dark Matter from Gamma-Ray Observations of the Coma Cluster of Galaxies with VERITAS and \Fermi

### 2012-05-31 ms-rev.tex
revision of CR part finished

\begin{abstract}
Observations of radio halos and relics in galaxy clusters indicate efficient electron
acceleration. Protons should likewise be accelerated and, on account of
weak energy losses, can accumulate, suggesting that
clusters may also be sources of very high-energy (VHE; $E>100$ GeV) gamma-ray
emission. We
report here on VHE gamma-ray observations of the Coma galaxy cluster
with the VERITAS array of
imaging Cherenkov telescopes, with complementing \Fermi-LAT observations at
GeV energies. No significant gamma-ray emission from the Coma cluster was detected. Integral flux
upper limits at the 99\% confidence level were measured to be on the
order of $(2-5)\times 10^{-8}\ {\rm ph.\,m^{-2}\,s^{-1}}$ (VERITAS, $>220\ {\rm GeV}$) and $\sim 3\times 10^{-6}\ {\rm ph.\,m^{-2}\, s^{-1}}$
(\Fermi, $1-3\ {\rm GeV}$), respectively. We use the gamma-ray upper
limits to constrain CRs and magnetic fields in Coma. Using an analytical
approach, the CR-to-thermal pressure ratio is
constrained to be $< 16\%$ from VERITAS data and $< 2.5\%$ from \Fermi data (averaged within the
virial radius). Assuming that the radio-emitting electrons of the Coma halo result from hadronic
CR interactions, the observations imply a lower limit on the central magnetic field in Coma of
$1-3\,\mu{\rm G}$, depending on the radial magnetic-field profile and on the gamma-ray
spectral index. Since these values are well below those inferred by Faraday rotation measurements
in Coma, this makes the hadronic model a very plausible explanation of the Coma radio
halo. Finally, since galaxy clusters are dark-matter (DM) dominated, the VERITAS upper limits have
been used to place constraints on the thermally-averaged product of the total 
self-annihilation
cross section and the relative velocity of the DM particles, $\expval{\sigma v}$.
\end{abstract}

@@

While several observations of clusters of galaxies have been made with satellite-borne and
ground-based gamma-ray telescopes, a detection of gamma-ray emission from a cluster has yet to be
made. Observations with EGRET \citep{article:Sreekumar_etal:1996, article:Reimer_etal:2003} and the
\Fermi Gamma-ray Space Telescope \citep{article:Ackermann_etal:2010} have provided upper limits on
the gamma-ray fluxes (typically $\sim10^{-9}$ ph cm$^{2}$ s$^{-1}$ for \Fermi observations) for
several galaxy clusters in the MeV to GeV band. Upper limits on the very-high-energy (VHE)
gamma-ray flux from a small sample of clusters, including the Coma cluster, have been provided by
observations with ground-based imaging atmospheric Cherenkov telescopes
\citep[IACTs;][]{article:Perkins_etal:2006, inproc:Perkins_etal:2008, article:Aharonian_etal:2009a,
article:Aleksic_etal:2010}.

@@

Table \ref{table:constraints_simple} shows the resulting constraints on the CR-to-thermal
pressure ratio, $X_{\CR} = \expval{P_{\CR}}/\expval{P_\rmn{th}}$, averaged within the virial radius,
$R_\rmn{vir}=2.2$~Mpc, that we define as the radius of a sphere enclosing a mean density that is 200
times the critical density of the Universe. Constraints on $X_\CR$ with 
VERITAS flux upper limits (99\% CL) strongly depend on $\alpha$.
This is due to the comparably large energy range from GeV
energies (that dominate the CR pressure, provided $\alpha>2$ and the CR population has a
nonrelativistic low-momentum cutoff, i.e., $q<m_{p}c$, where $m_{p}$ is the proton mass) to energies
at 220 GeV, where our quality selection criteria imposed the energy threshold. These gamma-ray
energies correspond to 1.6 TeV CRs -- an energy ratio of more than 3 orders of magnitude, which
explains the sensitivity to small changes in $\alpha$. The flux measurements within 0.2$^{\circ}$
are the most constraining due to a competition between the integrated signal and the background as
the integration radius increases.  This yields limits on $X_\CR$ between 0.048 and 0.43 (for
$\alpha$ varying between 2.1 and 2.5), with a constraint of $X_\CR<0.1$ for $\alpha=2.3$ (close to
the spectral index predicted by the simulations of \citet{article:PinzkePfrommer:2010} around 220
GeV). Constraints on $X_\CR$ with \Fermi limits (99\%\ CL) depend only weakly on $\alpha$ because GeV-band gamma rays are produced by CRs with energies near the relativistic transition, that dominantly contribute to the CR pressure.
$X_\CR$-constraints with \Fermi limits are most constraining for an aperture of
0.4$^{\circ}$; despite the slightly weaker flux upper limits in comparison to the smaller radii of
integration, we expect a considerably larger gamma-ray luminosity due to the increasing volume in
this model. The best limit of $X_\CR< 0.017$ is achieved for $\alpha=2.3$, while the limit for
$\alpha=2.1$ is only slightly worse $(X_\CR<0.025)$.

@@

These cosmological simulations only consider advective transport of CRs by turbulent gas motions
which lead to centrally-enhanced density profiles. However, other means of CR transport such as
diffusion and streaming may flatten the CR radial profiles. It has been suggested that advection is
dominant only during mergers, in which case relaxed clusters could have a gamma-ray luminosity 
reduced by up to a factor of five compared with merging clusters
\citep{article:EnsslinPfrommerMiniatiSubramanian:2011}. Hence, tight upper limits on the gamma-ray
emission can constrain a combination of acceleration physics and transport properties of CRs.

We adopt the density profile of thermal electrons as discussed in \S \ref{sec:simple} and model the
temperature profile
of the Coma cluster with a constant central temperature of $kT= 8.25$~keV and a characteristic
decline toward the cluster periphery in accordance with a fit to the universal profile derived from
cosmological cluster simulations \citep{article:PinzkePfrommer:2010, article:Pfrommer_etal:2007}
and the behavior of a nearby sample of deep {\em Chandra} cluster data
\citep{article:Vikhlinin_etal:2005}. This enables us to adopt the spatial and spectral distribution
of CRs according to the model by \citet{article:PinzkePfrommer:2010} that neglects the contribution of supernova remnants, AGN,
of cluster galaxies.

Figure \ref{fig:spectrum} shows the expected integral spectral energy distribution of Coma within
the virial radius (dotted line). This suggests a spectral index of $\alpha=2.1$ in the energy
interval 1-3 GeV and $\alpha=2.3$ for energies probed by VERITAS ($>220$~GeV). Also
shown are integrals of the differential spectrum for finite energy intervals across the angular
apertures tested in this study (solid lines). These model fluxes (summarized in Table
\ref{table:constraints_simple}) are compared to \Fermi and VERITAS flux upper limits for the same
energy intervals. Constraints on $X_\CR$ with the gamma-ray flux limit of \Fermi in the energy
interval 1-3 GeV ($<0.4^\circ$) are most constraining, but are still above the model predictions
that assume an optimistically large shock-acceleration efficiency and CR transport parameters as
laid out above. In our further analysis, we use the most constraining \Fermi flux limits in the
energy interval 1-3 GeV as well as the gamma-ray flux limits of VERITAS in the energy range above
220 GeV.

Figure \ref{fig:xcr} shows the CR-to-thermal pressure ratio, $X_{\CR} =
\expval{P_{\CR}}/\expval{P_\rmn{th}}$, as a function of radial distance, $R$, from the Coma cluster
center and contained within $R$. All radii are shown in units of the virial radius,
$R_\rmn{vir}=2.2$~Mpc. To compute the CR pressure, we assume a low-momentum cutoff of the CR
distribution at $q = 0.8\,m_{p}c$, where $m_{p}$ is the proton mass, to reflect the high cooling
rates at low CR energies. The CR-to-thermal pressure ratio rises toward the outer regions
on account of the higher efficiency of CR acceleration at the peripheral accretion shocks compared
to the weak central flow shocks. Adiabatic compression of a mixture of CRs and thermal gas disfavors
the CR pressure relative to the thermal pressure on account of the softer equation of state of
CRs. The weak increase of $X_{\CR}$ toward the core is due to the comparably fast thermal cooling of
gas.

@@

Equivalently, the most constraining \Fermi upper limit in the energy interval 1-3 GeV ($<0.4^\circ$)
is a factor of 1.1 larger than our model predictions (assuming $\alpha=2.1$ which is very close to
the simulated spectral index for the energy range 1-3 GeV). Scaling our integrated CR-to-thermal
pressure profile yields a constraint of $X_{\CR}<0.018$ within 0.4$^{\circ}$ that translates to a
limit within the cluster virial radius of $X_\CR<0.024$ (dashed lines of Figure \ref{fig:xcr}). The
$X_\CR$ constraint evaluated within the cluster virial radius is comparable to the constraint of
$X_\CR<0.025$ in our simplified model. Naturally, with the \Fermi limits we probe the region around
GeV energies that dominate the CR pressure, and we do not expect any differences in comparison to
the simplified power-law model in comparison to the universal CR spectrum with its concave CR
spectrum found in the simulations.

@@

For clusters that host radio halos, we can derive a minimum gamma-ray flux in the hadronic model of
radio halos -- where the radio-emitting electrons are secondaries from CR interactions.  Hadronic
interactions channel about the same power into secondary electrons and $\pi^{0}$-decay gamma rays. A
stationary distribution of CR electrons loses all its energy to synchrotron radiation for strong
magnetic fields ($B \gg B_\rmn{CMB} \simeq (1+z)^2\,(3.2 \mu{\rm G})$, 
where $B_\rmn{CMB}$ is the equivalent magnetic
field strength of the CMB so that $B_\rmn{CMB}^2/8\pi$ equals the CMB energy density). Thus the
ratio of gamma-ray to synchrotron flux becomes independent of the spatial distribution of CRs and
thermal gas \citep{article:Voelk:1989, article:Pohl:1994, article:Pfrommer:2008}, in particular with
$\alpha_{\nu}\simeq 1$ as the observed synchrotron spectral index.  Hence we can derive a minimum
gamma-ray flux in the hadronic model
\begin{equation}
\label{eq:Fmin}
F_{\gamma,\rmn{min}} = \frac{\dps A_{\gamma}}{\dps A_{\nu}}\frac{\dps L_{\nu}}{\dps 4\pi D_{\rmn{lum}}^{2}},
\end{equation}
where $L_{\nu}$ is the observed luminosity of the radio mini-halo, $D_{\rmn{lum}}$ denotes the
luminosity distance to the respective cluster, and $A_\gamma$ and $A_\nu$ are dimensional constants
that depend on the hadronic physics of the interaction \citep{article:Pfrommer:2008,
Pfrommer_etal:2008}. Lowering the magnetic field would require an increase in the energy density of
CR electrons to reproduce the observed synchrotron luminosity and thus increase the associated
gamma-ray flux.

@@

\subsection{Constraining the Magnetic Field}
\label{sec:B}
In the previous section, we have obtained an absolute lower limit on the gamma-ray emission in
the hadronic model by assuming high magnetic fields, $B\gg B_\rmn{CMB}$. We can turn the
argument around and use our upper limit on the gamma-ray emission (and by extension on the
CR pressure) to infer a lower limit on the magnetic field needed to explain the observed radio
emission. This, again, assumes the validity of the hadronic model of radio halos, in which the
radio-emitting electrons are secondaries from CR interactions. A stronger gamma-ray
constraint will tighten the magnetic-field limit. In case of a conflict with magnetic field
measurements by other methods, e.g., Faraday rotation measure (RM),\footnote{Generally,
Faraday RM analyses of the magnetic field strength by, e.g., background sources observed through
clusters, are degenerate with the magnetic coherence scale and may be biased by the unknown
correlation between magnetic and density fluctuations.} the hadronic model of radio halos would
be challenged. The method we use to constrain the magnetic field inherits a dependence on
the assumed radial scaling which we parametrize as
\begin{equation}
\label{eq:B}
B(r) = B_{0} \,\left(\frac{n_{\rmn{e}}(r)}{n_{\rmn{e}}(0)}\right)^{\alpha_B},
\end{equation}
as suggested by Faraday RM studies and numerical magnetohydrodynamical (MHD) simulations
\citep[][and references therein]{article:Bonafede_etal:2010, article:Bonafede_etal:2011}. Here
$n_{\rmn{e}}$ denotes the Coma electron density profile \citep{article:BrielHenryBohringer:1992}.
In fact, the magnetic field in the Coma cluster is among the best constrained, because its proximity
permits RM observations of seven radio sources located at projected distances of 50 to 1500 kpc from
the cluster center. The best-fit model yields $B_{0} = 4.7^{+0.7}_{-0.8}\,\mu$G and $\alpha_{B} =
0.5^{+0.2}_{-0.1}$ \citep{article:Bonafede_etal:2010}. We aim to constrain the central field
strength, $B_{0}$, and we permit the magnetic decline, $\alpha_{B}$, to vary within a reasonable
range of $\Delta\alpha_{B}=0.2$ as suggested by the Faraday RM studies. We proceed as follows:

@@

Table \ref{table:Bmin} shows the resulting lower limit of the central magnetic field ranging from
$B_{0} = 0.5$ to $1.4\,\mu$G in case the of VERITAS and from $B_{0} = 1.1$ to $4.6\,\mu$G in the
case of \Fermi.\footnote{Note that a central magnetic field of $3\,\mu$G corresponds in the Coma
  cluster to a magnetic-to-thermal energy density ratio of $X_B=0.005$.} Since these lower limits on
$B_{0}$ are below the values favored by Faraday RM for the entire parameter space spanned by
$\alpha_{B}$ and $\alpha$, the hadronic model is a viable explanation of the Coma radio halo. Future
gamma-ray observations of the Coma cluster may put more stringent constraints on the parameters of
the hadronic model.

@@

As a final word, in Table~\ref{table:Bmin} we additionally show the corresponding values for the
CR-to-thermal pressure ratio (at the largest emission radius at 1~Mpc) such that the model
reproduces the observed radio surface-brightness profile.\footnote{Note that in this section, we
determine the radial behavior of $X_\CR$ by adopting a specific model for the magnetic field and
requiring the modeled synchrotron surface-brightness profile to match the observed data of the
Coma radio halo. This is in contrast to the simplified analytical CR model where $X_\CR$ is
constant (\S~\ref{sec:simple}) and to the simulation-based model where $X_\CR(r)$ is derived from
cosmological cluster simulations (\S~\ref{sec:simulation}).} They should be interpreted as upper
limits since they are derived from flux upper limits. For the \Fermi upper limits, they range from
0.1 to 0.4; hence the $X_{\CR}$ profiles always obey the energy condition, i.e., $P_{\CR} <
P_{\mathrm{th}}$, over the entire range of the radio-halo emission ($< 1$ Mpc).\footnote{See Figure
  3 in \citet{article:PfrommerEnsslin:2004a} for the entire parameter range assuming minimum energy
  conditions, and \citet{article:PfrommerEnsslin:2004b}, Figure 7 for a parametrization as adopted
  in this study. We caution, however, that the minimum-energy condition is violated at the outer
  radio-halo boundary for the range of minimum magnetic-field values inferred by this study.} The
corresponding values for $X_\CR$ in the cluster center are smaller than 0.01 for the entire
parameter space probed in this study. We conclude that the hadronic model is not challenged by
current Faraday RM data and is a perfectly viable possibility in explaining the Coma radio-halo
emission.

@@

\section{Emission from Dark Matter Annihilations}
As already mentioned in the introduction, most of the mass in a galaxy cluster is in the form of
DM. While the nature of DM remains unknown, the most compelling theoretical candidate is a
WIMP. The self-annihilation of WIMPs can produce either monoenergetic gamma-ray lines or a
continuum of secondary gamma rays that deviates significantly from the power-law spectra observed
from most conventional astrophysical sources, with a sharp cut-off at the WIMP mass. These spectral
features together with the expected difference in the intensity distribution compared to
conventional astrophysical sources allow a clear, indirect detection of DM.

The expected gamma-ray flux due to self-annihilation of WIMPs in a
dark-matter halo is given by \citep{2004PhRvD..69l3501E,2008ApJ...678..594W,2010ApJ...720.1174A}
\begin{equation}
\frac{d\Phi_{\gamma}(\Delta\Omega,E)}{dE}=
\frac{\expval{\sigma v}}{8\pi m_{\chi}^{2}}\,\frac{dN_{\gamma}}{dE}\, J(\Delta\Omega),
\label{eqn:WIMPflux}
\end{equation}
where $\expval{\sigma v}$ is the thermally-averaged product of the total self-annihilation cross
section and the relative WIMP velocity, $m_{\chi}$ is the WIMP mass, $\frac{dN_{\gamma}}{dE}$ is the
differential gamma-ray yield per annihilation\footnote{In this work, we have calculated the
differential gamma-ray yield per annihilation using the Pythia Monte Carlo simulator.},
$\Delta\Omega$ is the observed solid angle, and $J$ is the so-called
astrophysical factor -- a 
factor which determines the DM annihilation rate and depends on the DM distribution.

>>>>

\begin{abstract}
  Observations of radio halos and relics in galaxy clusters indicate efficient electron
  acceleration. Protons should likewise be accelerated and, on account of weak energy losses, can
  accumulate, suggesting that clusters may also be sources of very high-energy (VHE; $E>100$ GeV)
  gamma-ray emission. We report here on VHE gamma-ray observations of the Coma galaxy cluster with
  the VERITAS array of imaging Cherenkov telescopes, with complementing \Fermi-LAT observations at
  GeV energies. No significant gamma-ray emission from the Coma cluster was detected. Integral flux
  upper limits at the 99\% confidence level were measured to be on the order of $(2-5)\times
  10^{-8}\ {\rm ph.\,m^{-2}\,s^{-1}}$ (VERITAS, $>220\ {\rm GeV}$) and $\sim 2\times 10^{-6}\ {\rm
    ph.\,m^{-2}\, s^{-1}}$ (\Fermi, $1-3\ {\rm GeV}$), respectively. We use the gamma-ray upper
  limits to constrain CRs and magnetic fields in Coma. Using an analytical approach, the
  CR-to-thermal pressure ratio is constrained to be $< 16\%$ from VERITAS data and $< 1.7\%$ from
  \Fermi data (averaged within the virial radius). {\bf These upper limits are now starting to
    constrain the CR physics in self-consistent cosmological cluster simulations and cap the maximum
    CR acceleration efficiency at structure formation shocks to be $<50\%$. Alternatively, this may
    argue for non-negligible CR transport processes such as CR streaming and diffusion into the
    outer cluster regions. } Assuming that the radio-emitting electrons of the Coma halo result from
  hadronic CR interactions, the observations imply a lower limit on the central magnetic field in
  Coma of $\sim (2 - 5.5)\,\mu{\rm G}$, depending on the radial magnetic-field profile and on the
  gamma-ray spectral index. Since these values are {\bf (for most of the parameter space)} below
  those inferred by Faraday rotation measurements in Coma, this {\bf renders} the hadronic model a
  very plausible explanation of the Coma radio halo. Finally, since galaxy clusters are dark-matter
  (DM) dominated, the VERITAS upper limits have been used to place constraints on the
  thermally-averaged product of the total self-annihilation cross section and the relative velocity
  of the DM particles, $\expval{\sigma v}$.
\end{abstract}

@@

While several observations of clusters of galaxies have been made with satellite-borne and
ground-based gamma-ray telescopes, a detection of gamma-ray emission from a cluster has yet to be
made. Observations with EGRET \citep{article:Sreekumar_etal:1996, article:Reimer_etal:2003} and the
\Fermi Gamma-ray Space Telescope \citep{article:Ackermann_etal:2010} have provided upper limits on
the gamma-ray fluxes (typically $\sim10^{-9}$ ph cm$^{2}$ s$^{-1}$ for \Fermi observations) for
several galaxy clusters in the MeV to GeV band. Upper limits on the very-high-energy (VHE)
gamma-ray flux from a small sample of clusters, including the Coma cluster, have been provided by
observations with ground-based imaging atmospheric Cherenkov telescopes
\citep[IACTs;][]{article:Perkins_etal:2006, inproc:Perkins_etal:2008, article:Aharonian_etal:2009a,
article:Aleksic_etal:2010,article:Aleksic_etal:2012}.

@@

Table \ref{table:constraints_simple} shows the resulting constraints on the CR-to-thermal
pressure ratio, $X_{\CR} = \expval{P_{\CR}}/\expval{P_\rmn{th}}$, averaged within the virial radius,
$R_\rmn{vir}=2.2$~Mpc, that we define as the radius of a sphere enclosing a mean density that is 200
times the critical density of the Universe. Constraints on $X_\CR$ with 
VERITAS flux upper limits (99\% CL) strongly depend on $\alpha$.
This is due to the comparably large energy range from GeV
energies (that dominate the CR pressure, provided $\alpha>2$ and the CR population has a
nonrelativistic low-momentum cutoff, i.e., $q<m_{p}c$, where $m_{p}$ is the proton mass) to energies
at 220 GeV, where our quality selection criteria imposed the energy threshold. These gamma-ray
energies correspond to 1.6 TeV CRs -- an energy ratio of more than 3 orders of magnitude, which
explains the sensitivity to small changes in $\alpha$. The flux measurements within 0.2$^{\circ}$
are the most constraining due to a competition between the integrated signal and the background as
the integration radius increases.  This yields limits on $X_\CR$ between 0.048 and 0.43 (for
$\alpha$ varying between 2.1 and 2.5), with a constraint of $X_\CR<0.1$ for $\alpha=2.3$ (close to
the spectral index predicted by the simulations of \citet{article:PinzkePfrommer:2010} around 220
GeV). Constraints on $X_\CR$ with \Fermi limits (99\%\ CL) depend only weakly on $\alpha$ because GeV-band gamma rays are produced by CRs with energies near the relativistic transition, that dominantly contribute to the CR pressure.
$X_\CR$-constraints with \Fermi limits are most constraining for an aperture of
0.4$^{\circ}$; despite the slightly weaker flux upper limits in comparison to the smaller radii of
integration, we expect a considerably larger gamma-ray luminosity due to the increasing volume in
this model. The best limit of $X_\CR< 0.012$ is achieved for $\alpha=2.3$, while the limit for
$\alpha=2.1$ is only slightly worse $(X_\CR<0.017)$.

@@

These cosmological simulations only consider advective transport of CRs by {\bf bulk gas flows that
  inject a turbulent cascade, leading to} centrally-enhanced density profiles. However, other means
of CR transport such as diffusion and streaming may flatten the CR radial profiles. {\bf The CRs
  stream along magnetic field lines in the opposite direction of the CR number density gradient (at
  any energy). In the stratified cluster atmosphere, this implies a net flux of CRs towards larger
  radii, equalizing the CR number density with time if not counteracted by advective transport. It
  has been suggested that advection velocities only dominate over the CR streaming velocities for
  periods with trans- and supersonic cluster turbulence during a cluster merger and drop below the
  CR streaming velocities for relaxing clusters.  As a consequence, a bimodality of the CR spatial
  distribution is expected to result; with merging (relaxed) clusters showing a centrally
  concentrated (flat) CR energy density profile
  \citep{article:EnsslinPfrommerMiniatiSubramanian:2011}. This translates into a bimodality of the
  expected diffuse radio and gamma-ray emission of clusters, since more centrally concentrated CRs
  will find higher target densities for hadronic CR proton interactions. As a result of this,
  relaxed clusters could have a reduced gamma-ray luminosity by up to a factor of five
  \citep{article:EnsslinPfrommerMiniatiSubramanian:2011}.}  Hence, tight upper limits on the
gamma-ray emission can constrain a combination of acceleration physics and transport properties of
CRs.

We adopt the density profile of thermal electrons as discussed in \S \ref{sec:simple} and model the
temperature profile
of the Coma cluster with a constant central temperature of $kT= 8.25$~keV and a characteristic
decline toward the cluster periphery in accordance with a fit to the universal profile derived from
cosmological cluster simulations \citep{article:PinzkePfrommer:2010, article:Pfrommer_etal:2007}
and the behavior of a nearby sample of deep {\em Chandra} cluster data
\citep{article:Vikhlinin_etal:2005}. This enables us to adopt the spatial and spectral distribution
of CRs according to the model by \citet{article:PinzkePfrommer:2010} that neglects the contribution of supernova remnants, AGN,
and cluster galaxies.

Figure \ref{fig:spectrum} shows the expected integral spectral energy distribution of Coma within
the virial radius (dotted line). This suggests a spectral index of $\alpha=2.1$ in the energy
interval 1-3 GeV and $\alpha=2.3$ for energies probed by VERITAS ($>220$~GeV). Also shown are
integrals of the differential spectrum for finite energy intervals across the angular apertures
tested in this study ({\bf dashed} lines). These model fluxes (summarized in Table
\ref{table:constraints_simple}) are compared to \Fermi and VERITAS flux upper limits for the same
energy intervals. Constraints on $X_\CR$ with the gamma-ray flux limit of \Fermi in the energy
interval 1-3 GeV ($<0.4^\circ$) are most constraining, {\bf since that combination of a specific
  energy interval and aperture minimizes the ratio of the upper limit to the expected model flux. In
  particular, this upper limit is 24\% below} the model predictions that assume an optimistically
large shock-acceleration efficiency and CR transport parameters as laid out above. {\bf Hence this
  enables us to constrain a combination of maximum shock acceleration efficiency and CR transport
  parameters.}  In our further analysis, we use the most constraining \Fermi flux limits in the
energy interval 1-3 GeV as well as the gamma-ray flux limits of VERITAS in the energy range above
220 GeV.

Figure \ref{fig:xcr} shows the CR-to-thermal pressure ratio, $X_{\CR} =
\expval{P_{\CR}}/\expval{P_\rmn{th}}$, as a function of radial distance, $R$, from the Coma cluster
center and contained within $R$. All radii are shown in units of the virial radius,
$R_\rmn{vir}=2.2$~Mpc. To compute the CR pressure, we assume a low-momentum cutoff of the CR
distribution at $q = 0.8\,m_{p}c$, where $m_{p}$ is the proton mass. {\bf This is suggested by
  cosmological cluster simulations and reflects} the high {\bf Coulomb} cooling rates at low CR
energies. The CR-to-thermal pressure ratio rises toward the outer regions on account of the higher
efficiency of CR acceleration at the peripheral accretion shocks compared to the weak central flow
shocks. Adiabatic compression of a mixture of CRs and thermal gas disfavors the CR pressure relative
to the thermal pressure on account of the softer equation of state of CRs. The weak increase of
$X_{\CR}$ toward the core is due to the comparably fast thermal cooling of gas.

@@

{\bf As already eluded,} the most constraining \Fermi upper limit in the energy interval 1-3 GeV
($<0.4^\circ$) is a factor of {\bf 0.76 smaller} than our model predictions (assuming $\alpha=2.1$
which is very close to the simulated spectral index for the energy range 1-3 GeV). Scaling our
integrated CR-to-thermal pressure profile yields a constraint of $X_{\CR}<0.012$ within
0.4$^{\circ}$ that translates to a limit within the cluster virial radius of $X_\CR<0.017$ (dashed
lines of Figure \ref{fig:xcr}). The $X_\CR$ constraint evaluated within the cluster virial radius is
comparable to the constraint of $X_\CR<0.017$ in our simplified model. Naturally, with the \Fermi
limits we probe the region around GeV energies that dominate the CR pressure, and we do not expect
any differences in comparison to the simplified power-law model in comparison to the universal CR
spectrum with its concave CR spectrum found in the simulations.

@@

For clusters that host radio halos, we can derive a minimum gamma-ray flux in the hadronic model of
radio halos -- where the radio-emitting electrons are secondaries from CR interactions.  Hadronic
interactions channel about the same power into secondary electrons and $\pi^{0}$-decay gamma rays. A
stationary distribution of CR electrons loses all its energy to synchrotron radiation for strong
magnetic fields ($B \gg B_\rmn{CMB} \simeq (1+z)^2\,3.2 \mu{\rm G}$, 
where $B_\rmn{CMB}$ is the equivalent magnetic
field strength of the CMB so that $B_\rmn{CMB}^2/8\pi$ equals the CMB energy density). Thus the
ratio of gamma-ray to synchrotron flux becomes independent of the spatial distribution of CRs and
thermal gas \citep{article:Voelk:1989, article:Pohl:1994, article:Pfrommer:2008}, in particular with
$\alpha_{\nu}\simeq 1$ as the observed synchrotron spectral index.  Hence we can derive a minimum
gamma-ray flux in the hadronic model
\begin{equation}
\label{eq:Fmin}
F_{\gamma,\rmn{min}} = \frac{\dps A_{\gamma}}{\dps A_{\nu}}\frac{\dps L_{\nu}}{\dps 4\pi D_{\rmn{lum}}^{2}},
\end{equation}
where $L_{\nu}$ is the observed luminosity of the radio mini-halo, $D_{\rmn{lum}}$ denotes the
luminosity distance to the respective cluster, and $A_\gamma$ and $A_\nu$ are dimensional constants
that depend on the hadronic physics of the interaction \citep{article:Pfrommer:2008,
Pfrommer_etal:2008}. Lowering the magnetic field would require an increase in the energy density of
CR electrons to reproduce the observed synchrotron luminosity and thus increase the associated
gamma-ray flux.

@@

\subsection{Constraining the Magnetic Field}
\label{sec:B}
In the previous section, we have obtained an absolute lower limit on the gamma-ray emission in
the hadronic model by assuming high magnetic fields, $B\gg B_\rmn{CMB}$. We can turn the
argument around and use our upper limit on the gamma-ray emission (and by extension on the
CR pressure) to infer a lower limit on the magnetic field needed to explain the observed radio
emission. This, again, assumes the validity of the hadronic model of radio halos, in which the
radio-emitting electrons are secondaries from CR interactions. A stronger gamma-ray
constraint will tighten the magnetic-field limit. In case of a conflict with magnetic field
measurements by other methods, e.g., Faraday rotation measure (RM),\footnote{Generally,
Faraday RM analyses of the magnetic field strength by, e.g., background sources observed through
clusters, are degenerate with the magnetic coherence scale and may be biased by the unknown
correlation between magnetic and density fluctuations.} the hadronic model of radio halos would
be challenged. The method we use to constrain the magnetic field inherits a dependence on
the assumed radial scaling which we parametrize as
\begin{equation}
\label{eq:B}
B(r) = B_{0} \,\left(\frac{n_{\rmn{e}}(r)}{n_{\rmn{e}}(0)}\right)^{\alpha_B},
\end{equation}
as suggested by Faraday RM studies and numerical magnetohydrodynamical (MHD) simulations
\citep[][and references therein]{article:Bonafede_etal:2010, article:Bonafede_etal:2011}. Here
$n_{\rmn{e}}$ denotes the Coma electron density profile \citep{article:BrielHenryBohringer:1992}.
In fact, the magnetic field in the Coma cluster is among the best constrained, because its proximity
permits RM observations of seven radio sources located at projected distances of 50 to 1500 kpc from
the cluster center. The best-fit model yields $B_{0} = 4.7^{+0.7}_{-0.8}\,\mu$G and $\alpha_{B} =
0.5^{+0.2}_{-0.1}$ \citep{article:Bonafede_etal:2010}. We aim to constrain the central field
strength, $B_{0}$, and we permit the magnetic decline, $\alpha_{B}$, to vary within a reasonable
range of $\Delta\alpha_{B}=0.2$ as suggested by {\bf those} Faraday RM studies. We proceed as follows:

@@

Table \ref{table:Bmin} shows the resulting lower limit of the central magnetic field ranging from
$B_{0} = 0.5$ to $1.4\,\mu$G in case the of VERITAS and from $B_{0} = 1.4$ to $5.5\,\mu$G in the
case of \Fermi.\footnote{Note that a central magnetic field of $3\,\mu$G corresponds in the Coma
  cluster to a magnetic-to-thermal energy density ratio of $X_B=0.005$.} Since these lower limits on
$B_{0}$ are below the values favored by Faraday RM for {\bf most of the} parameter space spanned by
$\alpha_{B}$ and $\alpha$ {\bf (and never exceed the values for the phenomenological Faraday
  RM-inferred $B$-model)}, the hadronic model is a viable explanation of the Coma radio halo. {\bf
  In fact, the \Fermi upper limits start to rule out the parameter combination of
  $\alpha_{B}\gtrsim 0.7$ and $\alpha \gtrsim 2.5$ for the hadronic model of the Coma radio halo.}
Future gamma-ray observations of the Coma cluster may put more stringent constraints on the
parameters of the hadronic model.

@@

As a final word, in Table~\ref{table:Bmin} we additionally show the corresponding values for the
CR-to-thermal pressure ratio (at the largest emission radius at 1~Mpc) such that the model
reproduces the observed radio surface-brightness profile.\footnote{Note that in this section, we
determine the radial behavior of $X_\CR$ by adopting a specific model for the magnetic field and
requiring the modeled synchrotron surface-brightness profile to match the observed data of the
Coma radio halo. This is in contrast to the simplified analytical CR model where $X_\CR$ is
constant (\S~\ref{sec:simple}) and to the simulation-based model where $X_\CR(r)$ is derived from
cosmological cluster simulations (\S~\ref{sec:simulation}).} They should be interpreted as upper
limits since they are derived from flux upper limits. For the \Fermi upper limits, they range from
0.08 to 0.27; hence the $X_{\CR}$ profiles always obey the energy condition, i.e., $P_{\CR} <
P_{\mathrm{th}}$, over the entire range of the radio-halo emission ($< 1$ Mpc).\footnote{See Figure
  3 in \citet{article:PfrommerEnsslin:2004a} for the entire parameter range assuming minimum energy
  conditions, and \citet{article:PfrommerEnsslin:2004b}, Figure 7 for a parametrization as adopted
  in this study. We caution, however, that the minimum-energy condition is violated at the outer
  radio-halo boundary for the range of minimum magnetic-field values inferred by this study.} The
corresponding values for $X_\CR$ in the cluster center are smaller than 0.01 for the entire
parameter space probed in this study. We conclude that the hadronic model is not challenged by
current Faraday RM data and is a perfectly viable possibility in explaining the Coma radio-halo
emission.

@@

\section{Emission from Dark Matter Annihilations}
As already mentioned in the introduction, most of the mass in a galaxy cluster is in the form of
DM. While the nature of DM remains unknown, a compelling theoretical candidate is a
WIMP. The self-annihilation of WIMPs can produce either monoenergetic gamma-ray lines or a
continuum of secondary gamma rays that deviates significantly from the power-law spectra observed
from most conventional astrophysical sources, with a sharp cut-off at the WIMP mass. These spectral
features together with the expected difference in the intensity distribution compared to
conventional astrophysical sources allow a clear, indirect detection of DM.

The expected gamma-ray flux due to self-annihilation of WIMPs in a
dark-matter halo is given by 
\begin{equation}
\frac{d\Phi_{\gamma}(\Delta\Omega,E)}{dE}=
\frac{\expval{\sigma v}}{8\pi m_{\chi}^{2}}\,\frac{dN_{\gamma}}{dE}\, J(\Delta\Omega),
\label{eqn:WIMPflux}
\end{equation}
where $\expval{\sigma v}$ is the thermally-averaged product of the total self-annihilation cross
section and the relative WIMP velocity, $m_{\chi}$ is the WIMP mass, $\frac{dN_{\gamma}}{dE}$ is the
differential gamma-ray yield per annihilation\footnote{In this work, we have calculated the
differential gamma-ray yield per annihilation using the Pythia Monte Carlo simulator.},
$\Delta\Omega$ is the observed solid angle, and $J$ is the so-called
astrophysical factor -- a 
factor which determines the DM annihilation rate and depends on the DM distribution.
